---
title: "Today's Top Hits"
author: "Ankit Sanghi, Sarah Hussain"
output: html_document
runtime: shiny
resource_files:
- spotify_rforest_decade.rds
- spotify_rforest_popularity.rds
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(shiny)
library(readr)
library(caret)
library(randomForest)
all_data <- read_csv("./data.csv")
data_by_artist <- read_csv("./data_by_artist.csv")
data_by_year <- read_csv("./data_by_year.csv")
data_by_genres <- read_csv("./data_by_genres.csv")
data_w_genres <- read_csv("./data_w_genres.csv")
pivoted_data_by_year <- data_by_year %>%
        pivot_longer(cols = c("acousticness", 
                                 "danceability", 
                                 "energy", 
                                 "instrumentalness", 
                                 "valence", 
                                 "liveness", 
                                 "speechiness"),
                     names_to = "metric", 
                     values_to = "values")

model_results <- read_rds("./results.rds")
```


## Motivation and Background

Spotify grants its users access to over 60 million songs, ranging from over 3000 genres, 32000 artists, across 102 years. It also encodes certain descriptive metrics about the nature of each song. These include: danceability, acousticness, valency, and many more. This gives us the rare opportunity to see how these metrics change over time and whether there are relations between them. It also lets us analyze whether any given decade has a particular musical profile associated with it. How do musical profiles change across the decades? Does music overall get more popular over time? These are all questions that can be answered using our Spotify dataset, which is why we are conducting this analysis.

## Objectives

We wanted to answer four main questions. First, we wanted to see how song profiles change over the years. Does each decade have a distinct sound profile? Next, we wanted to look at how the popularity of music overall changes through the 20th and early 21st century. Does the overall popularity of music increase or decrease over time? Finally, given a song's profile, can we use statistical learning to predict how popular it would be and which decade it would fall in? These are all questions that could give us a lot of insight into how music has changed over the years and what kind of music remains popular across the ages.

## The Data

Our data was found on [Kaggle](https://www.kaggle.com/yamaerenay/spotify-dataset-19212020-160k-tracks). It was scraped by Yamac Eren Ay and contains 174389 songs with 19 variables. Each row is a song on Spotify identified by a unique track ID. We also used acousticness, danceability, energy,  instrumentalness, valence, and speechiness which all have a numeric range from 0 to 1. Additionally, we were also interested in the year in which the song was released as well as the popularity of the song which is measure on a scale of  0 to 100. This information can also be found on the Kaggle page for the data. Overall, the dataset is very robust and fully featured and can provide some very valuable insights into how music has evolved over the years.

## Analysis

### Song Profiles over Time
We use graphical analysis to observe how the descriptive metrics for each song over time. Below is a graph where you can toggle various metrics and compare them. 

```{r, echo=FALSE}
sidebarLayout(
    sidebarPanel(
      checkboxGroupInput("metrics", "Song Profile Metrics", 
                     choices = c("acousticness", 
                                 "danceability", 
                                 "energy", 
                                 "instrumentalness", 
                                 "valence", 
                                 "speechiness"),
                     selected = c("acousticness", "energy")),
      width = 3
    ),
    mainPanel(
      renderPlot({
        pivoted_data_by_year %>%
        filter(metric %in% input$metrics) %>%
        ggplot(aes(x = year, y = values, color = metric)) + geom_line()
      })
    )
  )
```

The graph below shows that the acousticness of songs decreases but the energy increases over time.  We see energy overtake acousticness in the 1960s, which may be attributed to the fact that instruments like the electric guitar started becoming more popular. More rock and pop artists became popular during this period such as The Beatles, The Rolling Stones, The Beach Boys, Janis Joplin, and many others. These genres of music tend to use more synthesized sounds which explains the drop in acousticness and rise in energy.

```{r, echo=FALSE, fig.align="center"}
pivoted_data_by_year %>%
        filter(metric %in% c("acousticness", "energy")) %>%
        ggplot(aes(x = year, y = values, color = metric)) + geom_line()
```

Next, let us look at the change in danceability and valence over time. Valence measures the musical positivness of a song. A low valence means the song sounds more negative and a high valence means it sounds more positive. If we look at the trend of valence and danceability, they are almost identical until the 2010s, where they start to diverge. It makes sense for them to be near identical as often music that is considered danceable is also positive sounding. So why do these two metrics diverge in the 2010s?  We observe that the energy of the songs increase along with danceability but they are also more negative sounding. This can be explained by the rise of rap music. Rap music often has a low beat that usually isn't categorized as sounding positive. The music used is often not cheerful or upbeat. Rather, it is heavy and serves as simply a background to the vocals. The rhythmic nature of rap makes it danceable and energetic, whereas the low sounds of the music gives it a low valence. Rap started picking up in the 2010s, where it beat rock and pop as the most popular genre of music. Thus, this could explain this divide between valence and danceability in the 2010s.

```{r, echo=FALSE, fig.align="center"}
pivoted_data_by_year %>%
        filter(metric %in% c("danceability", "valence")) %>%
        ggplot(aes(x = year, y = values, color = metric)) + geom_line()
```

Finally, let us look at instrumentalness and speechiness. Instrumentalness measures the amount of vocals in the song, and speechiness detects the presence of spoken words. We can see that until 1950, when one increases the other decreases. After this point though instrumentalness takes the lead and remains on top. While both metrics decrease over time we see an increase again in the early 2000s.

```{r, echo=FALSE, fig.align="center"}
pivoted_data_by_year %>%
        filter(metric %in% c("instrumentalness", "speechiness")) %>%
        ggplot(aes(x = year, y = values, color = metric)) + geom_line()
```

### Music Popularity Over Time

Our dataset also allows us to see the popularity of music from different periods of time. We can see in the graph below that songs before 1950 had popularity values of close to zero. Popularity begins to rise in the 1950s and continues to rise right up until 2000. Then it begins to sharply decline. This means that music made before 2000 is a lot more popular than the music after. This can be explained by the fact that it became much easier to upload music online and gain an audience. It also became easier to stream all kinds of music. Before the 2000s, there were only a few small groups that could make and publish music because of the large barrier to entry made by record labels and publishing companies. But after 2000, this barrier crumbled away and now anyone can upload their music online. This meant that instead of listeners focusing on a small group of bands, they now have a plethora of different genres and artists to choose from. Popularity of any given song after 2000 is lower, as there is a smaller concentration of listeners listening to a given song. Songs from before 2000 are also considered to be iconic because of the small number of bands that dominated the music industry (e.g. The Beatles, Queen, Nirvana). However, post 2000 because of the rise in smaller artists being able to publish music, people's attention was split and so popularity became distributed instead of being concentrated among a small number of bands. It's not that music after 2000 is not popular but rather, the popularity is distributed across all the songs due to music being easier to access and produce.

```{r, echo=FALSE, fig.align="center"}
data_by_year %>%
  ggplot(aes(x = year, y = popularity)) + geom_line()
```


## Which Musical Decade do you Really Belong In?

Predictions are made using a Random Forest Classification with 500 trees and the default mtry value (square-root of the number of predictors). Specifically, we use two prediction models. In the first one, the musical decade is the the main predictor and in the second one popularity of the song is the main predictor. The musical decades range from the 1920s to the 2020s and the popularity variable was grouped into 10 bins (0-10, 11-20, 21-30, 31-40, 41-50, 51-60, 61-70, 71-80, 81-90, 91-100). This was done primarily to reduces the number of classes. For both models we use the same set of predictors: acousticness, danceability, energy, instrumentalness, valence, and speechiness. 
 
The interactive feature below allows you to enter the metrics for music that you like to listen to. Our model will tell you which decade your musical tastes belong in as well as how popular a song with these metrics would be. The prediction will take a few seconds, so please be patient.
 
It is important to note that the accuracy of the models was 38% and 41% respectively. These are relatively low and may be attributed to the fact that there is a lot of heterogeneity in the type of music that is being listened to and so there is always more than one genre or song type that is popular. Music follows you through time and people tend to actively enjoy songs in the present that were also popular in the past. Thus, our random forest classification may not accurately represent which musical decade you really belong to. 

```{r, echo=FALSE}
shinyApp(
  ui <- fluidPage(
    sidebarLayout(
      sidebarPanel(
        sliderInput("acousticness", "How acoustic do you like your songs?", 
                    min = 0, 
                    max = 1, 
                    step = 0.1, 
                    value = 0.5),
          sliderInput("danceability", "How danceable do you like your songs?", 
                    min = 0, 
                    max = 1, 
                    step = 0.1, 
                    value = 0.5),
        sliderInput("energy", "How energetic do you like your songs?", 
                    min = 0, 
                    max = 1, 
                    step = 0.1, 
                    value = 0.5),
        sliderInput("instrumentalness", "How instrumental do you like your songs?", 
                    min = 0, 
                    max = 1, 
                    step = 0.1, 
                    value = 0.5),
        sliderInput("valence", "How positive do you like your songs?", 
                    min = 0, 
                    max = 1, 
                    step = 0.1, 
                    value = 0.5),
        sliderInput("speechiness", "How lyrical do you like your songs?", 
                    min = 0, 
                    max = 1, 
                    step = 0.1, 
                    value = 0.5),
        actionButton("go", "Predict")
      ),
    mainPanel(
      htmlOutput("result")
      )
    )
  ),
  server <- function(input, output) {
     observeEvent(input$go, {
      output_data <- model_results %>%
        filter(acousticness == input$acousticness, 
               danceability == input$danceability, 
               energy == input$energy, 
               instrumentalness == input$instrumentalness,
               valence == input$valence,
               speechiness == input$speechiness)
      
      output$result <- renderUI(HTML(
        paste("<h3>Your musical tastes belong in the ",
              output_data[["decade_results"]] ,"s and would have a popularity between ", 
              output_data[["popularity_result"]],"!</h3>", sep = "")))
    })
  },
  options = list(height = 900)
)
```




